### 2018-02-15 manuscript.tex
esteban's comments ++

\section*{Results}

@@

We gathered a subset of the data from a recent publication \citationneeded{}, which was gracefully made publicly available by the authors\citationneeded{zenodo}.
It shows that flies, which normally become arrhythmic in a Light-Light (LL) regime, are rhythmic (with an period greater than 24h\TODO{phrasing}) when the gene NKCCox is expressed in clock neurons.
The authors have recorded the activity of both TIM/NKCCOK (NKCCox is expressed in their clock neurons) and TIM/+ (the appropriate control) in LL for six days.
Prior to LL, the animals were recorded for a baseline three or four days in normal LD, 12h:12h, conditions.
Two repetitions were performed resulting in a total of 58 animals.
The \texttt{metadata.csv} file, which describes exhaustively all individuals, as well as all the associated result files can be downloaded on zenodo\citationneeded{ibid}.

@@

\subsection*{Preprocessing}
We notice, from the metadata, that the two replicates do not have the same baseline time.
We would like to express the time relative to the important event: the transition to \texttt{LL}. 
To do so, we subtract the \texttt{baseline\_days} metavariable from the \texttt{t} variable.
This gives us an opportunity to illustrate the use \texttt{xmv()}, which expands metavariables as variables.
In addition, we use the \texttt{data.table} syntax to create, in place, a \texttt{moving} variable.
It is \texttt{TRUE} when and only when \texttt{activity} is greater than zero:

@@

\begin{figure}[!h]
	%\includegraphics[width=1\textwidth]{fig/fig-3.pdf}
	\caption{{\bf Experiment quality control.}
			Tile plot representing the fraction of time spent moving as a colour intensity.
			Each individual is represented by a row and time, on the x-axis, is binned in 30 minutes.}
	\label{fig:fig-3}
\end{figure}

@@

The activity of dead or escaped animals is falsely scored as long series of zeros (see, for instance, individual 30 and 18 in Fig~\ref{fig:fig-3}A).
Our \texttt{sleepr} package offers a tool to detect and remove such artefactual data.
The updated version can be visualised in Fig~\ref{fig:fig-3}B.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# remove data after death}
\hlstd{dt} \hlkwb{<-} \hlstd{sleepr}\hlopt{::}\hlkwd{curate_dead_animals}\hlstd{(dt, moving)}
\hlcom{# same as above}
\hlstd{fig3B} \hlkwb{<-} \hlkwd{ggetho}\hlstd{(dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= label,} \hlkwc{z} \hlstd{= moving))} \hlopt{+}
    \hlkwd{stat_tile_etho}\hlstd{()} \hlopt{+}
    \hlkwd{stat_ld_annotations}\hlstd{(}\hlkwc{x_limits} \hlstd{=} \hlkwd{c}\hlstd{(dt[,} \hlkwd{min}\hlstd{(t)],} \hlnum{0}\hlstd{))} \hlopt{+}
    \hlkwd{stat_ld_annotations}\hlstd{(}\hlkwc{x_limits} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{, dt[,} \hlkwd{max}\hlstd{(t)]),}
                        \hlkwc{ld_colours} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"white"}\hlstd{,} \hlstr{"grey"}\hlstd{))}
\end{alltt}
\end{kframe}
\end{knitrout}

For the purpose of this example, we keep only individuals that have \emph{at least five days in LL}.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# for each id, we check for validity}
\hlstd{valid_dt} \hlkwb{<-} \hlstd{dt[,} \hlkwd{.}\hlstd{(}\hlkwc{valid} \hlstd{=} \hlkwd{max}\hlstd{(t)} \hlopt{>} \hlkwd{days}\hlstd{(}\hlnum{5}\hlstd{)),} \hlkwc{by} \hlstd{= id]}
\hlcom{# a vector of all valid ids}
\hlstd{valid_ids} \hlkwb{<-} \hlstd{valid_dt[valid} \hlopt{==} \hlstd{T, id]}
\hlcom{# filter dt with the valid ids}
\hlstd{dt} \hlkwb{<-} \hlstd{dt[id} \hlopt{%in%} \hlstd{valid_ids]}
\hlkwd{summary}\hlstd{(dt)}
\end{alltt}
\begin{verbatim}
## behavr table with:
##  52	individuals
##  9	metavariables
##  3	variables
##  1.40609e+05	measurements
##  1	key (id)
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\subsection*{Double-plotted actograms}
``Double-plotted actograms'' are a common visualisation of periodicity and rhythmicity in circadian experiments.
In  \nameref{S1-Fig}A, we show the double-plotted actograms of each animal. A representative sample of eight individuals is shown in Fig~\ref{fig:fig-4}A.

@@

\subsection*{Periodograms}
Ultimately, in order to quantify periodicity and rhythmicity, we compute periodograms.
Several methods are implemented in \texttt{zeitbebr}. In this example, we generate $\chi{}^2$ periodograms and lay them out in a grid.
A subset of eight animals is shown in Fig~\ref{fig:fig-4}B (see \nameref{S1-Fig}B for all individuals).

@@

\begin{figure}[!h]
	%\includegraphics[width=1\textwidth]{fig/fig-4.pdf}
	\caption{{\bf Visualisation of the periodicity in activity of eight representative animals.}
		A: Double plotted actograms showing activity over the experiment. Transition to LL happens at day 0.
		B: $\chi{}^2$ periodograms over the LL part of the experiment matching the animals in A.
		The blue annotation represents the first peak (if present) above the significance threshold (red line).
		Titles on top of each facet refer to the label allocated to each individual.
		A version of this figure with all animals is available \nameref{S1-Fig}.
	}
	\label{fig:fig-4}
\end{figure}

\subsection*{Population statistics}
Both double-plotted actograms and periodograms suggest that NKCCOX/+ flies are mostly 
arhythmic in LL whilst Tim/NKCCOX appear to have a consistent, long-period rhythm.
To visualise this at the population scale, we can plot an average periodogram (see Fig~\ref{fig:fig-5}A):

@@

\begin{figure}[!h]
	%\includegraphics[width=1\textwidth]{fig/fig-5.pdf}
	\caption{{\bf Population statistics on circadian phenotype.}
			A: Average periodograms. 
			      The aggregated relative power of the periodogram of all animals.
			      The solid lines and the shaded areas show population means and their 95\% bootstrap confidence interval, respectively.
			B: Peak periodicity power and average.
			      Values of the peak period for animals with a significant peak.
			      Individual animals are shown by dots whose size represent power of the peak period.
			      The error bars are 95\% bootstrap confidence interval on the population mean.
			C: Frequencies of rhythmic animals.
			      Number of rhythmic animals (\emph{i.e.} with a significant peak) in each genotypes.
			      Dark and clear fillings indicate rhythmic and arhythmic animals, respectively.
			      }
	\label{fig:fig-5}
\end{figure}

To further quantify this difference, we can show the value of the peak for both groups (see Fig~\ref{fig:fig-5}B). First of all, we compute a summary per individual (\texttt{by=id}):

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{summary_dt} \hlkwb{<-} \hlstd{per_dt[,}\hlkwd{.}\hlstd{(}
                        \hlkwc{first_peak_period} \hlstd{= period[peak}\hlopt{==}\hlnum{1}\hlstd{],}
                        \hlkwc{first_peak_power} \hlstd{= power[peak}\hlopt{==}\hlnum{1}\hlstd{],}
                        \hlkwc{is_rhythmic} \hlstd{=} \hlkwd{any}\hlstd{(peak}\hlopt{==}\hlnum{1}\hlstd{)),}
                      \hlkwc{by}\hlstd{=id]}

\hlcom{# rejoin metadata}
\hlstd{summary_dt} \hlkwb{<-} \hlkwd{rejoin}\hlstd{(summary_dt)}
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# standard ggplot}
\hlstd{fig5B} \hlkwb{<-} \hlkwd{ggplot}\hlstd{(summary_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= first_peak_period,}
                                \hlkwc{x} \hlstd{= genotype))} \hlopt{+}
              \hlcom{# draw the mean of each genotype group}
              \hlkwd{stat_summary}\hlstd{(}\hlkwc{fun.y} \hlstd{= mean,} \hlkwc{geom} \hlstd{=} \hlstr{"point"}\hlstd{,} \hlkwc{shape}\hlstd{=}\hlnum{3}\hlstd{)} \hlopt{+}
              \hlcom{# draw bootstrap confidence intervals}
              \hlkwd{stat_summary}\hlstd{(}\hlkwc{fun.data} \hlstd{= mean_cl_boot,} \hlkwc{geom} \hlstd{=} \hlstr{"errorbar"}\hlstd{)} \hlopt{+}
              \hlcom{# shows all individuals as points. }
              \hlcom{# the size of the point expresses the power of the peak}
              \hlkwd{geom_jitter}\hlstd{(}\hlkwd{aes}\hlstd{(}\hlkwc{colour}\hlstd{= genotype,}
                              \hlkwc{size}\hlstd{=first_peak_power),}
                          \hlkwc{alpha}\hlstd{=}\hlnum{.67}\hlstd{)} \hlopt{+}
              \hlcom{# We would like to convert time in hour}
              \hlkwd{scale_y_hours}\hlstd{(}\hlstr{"Period"}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

To convey that Tim/NKCCOX mutants are more rhythmic, we could represent the proportion of rhythmic flies (see Fig~\ref{fig:fig-5}C):
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# standard ggplot}
\hlstd{fig5C} \hlkwb{<-} \hlkwd{ggplot}\hlstd{(summary_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{x} \hlstd{= genotype,}
                                \hlkwc{fill} \hlstd{= genotype,}
                                \hlkwc{alpha} \hlstd{= is_rhythmic}
                                \hlstd{))} \hlopt{+}
              \hlkwd{geom_bar}\hlstd{(}\hlkwc{colour}\hlstd{=}\hlstr{"black"}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

\texttt{R} provides one of the richest statistical toolbox available, which allows users to go deeper tin he analysis of the extracted variables.
One could, for instance, perform a $\chi{}^2$ test on the number of rhythmic \emph{vs} arhythmic flies in both genotypes.
To address the same question, we fit a binomial generalised linear model, which shows a strong positive effect of genotype Tim/NKCCOX on the probability of being rhythmic ($p$-value $< 10^{-5}$):
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{fit} \hlkwb{<-} \hlkwd{glm}\hlstd{(is_rhythmic} \hlopt{~} \hlstd{genotype, summary_dt,} \hlkwc{family}\hlstd{=}\hlstr{"binomial"}\hlstd{)}

@@

Lastly, we can generate a table that compute arbitrary population statistics for each genotype:
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{result_dt} \hlkwb{<-} \hlstd{summary_dt[,}
          \hlkwd{.}\hlstd{(}
            \hlkwc{mean_period} \hlstd{=} \hlkwd{mean}\hlstd{(first_peak_period,} \hlkwc{na.rm} \hlstd{= T)} \hlopt{/} \hlkwd{hours}\hlstd{(}\hlnum{1}\hlstd{),}
            \hlkwc{sd_period} \hlstd{=} \hlkwd{sd}\hlstd{(first_peak_period,} \hlkwc{na.rm} \hlstd{= T)} \hlopt{/} \hlkwd{hours}\hlstd{(}\hlnum{1}\hlstd{),}
            \hlkwc{n_rhythmic} \hlstd{=} \hlkwd{sum}\hlstd{(is_rhythmic),}
            \hlkwc{n} \hlstd{= .N}
            \hlstd{),}
          \hlkwc{by} \hlstd{= genotype}
          \hlstd{]}
\hlstd{result_dt}
\end{alltt}
\begin{verbatim}
##      genotype mean_period sd_period n_rhythmic  n
## 1:   NKCCOX/+    25.92500  4.133098          4 22
## 2: Tim/NKCCOX    26.36071  1.834513         28 30
\end{verbatim}
\end{kframe}
\end{knitrout}

This examples shows how \texttt{rethomics} can be used from the raw data to making publication quality figures and statistics.
We were able to comprehensively analyse classical circadian data with a few line of code.
Since the interface is programmatic, it adapts really well to much larger data sets and provides a large degree of freedom and flexibility.

>>>>

\section*{Results}
\TODO{@esteban rewrite this please ;)}

@@

We gathered a subset of the data from a recent publication \citationneeded{}, which was gracefully made publicly available by the authors\citationneeded{zenodo}.
It shows that flies, which normally become arrhythmic in a Light-Light (LL) regime, are rhythmic (with an period greater than 24h\TODO{phrasing}) when the gene NKCCox is expressed in clock neurons.
The authors have recorded the activity of both TIM/NKCCOK (NKCCox is expressed in their clock neurons) and NKCCOK/+ (the appropriate control) in LL for six days.
Prior to LL, the animals were recorded for a baseline three or four days in normal LD, 12h:12h, conditions.
Two repetitions were performed resulting in a total of 58 animals.
The \texttt{metadata.csv} file, which describes exhaustively all individuals, as well as all the associated result files can be downloaded on zenodo\citationneeded{ibid}.

@@

\subsection*{Preprocessing}
The two replicates do not have the same baseline time and we would like to express the time relative to the important event:
the transition from LD to LL. 
Therefore, we subtract the \texttt{baseline\_days} metavariable from the \texttt{t} variable.
This gives us an opportunity to illustrate the use \texttt{xmv()}, which expands metavariables as variables.
In addition, we use the \texttt{data.table} syntax to create, in place, a \texttt{moving} variable.
It is \texttt{TRUE} when and only when \texttt{activity} is greater than zero:

@@

\begin{figure}[!h]
	%\includegraphics[width=1\textwidth]{fig/fig-3.pdf}
	\caption{{\bf Experiment quality control.}
			Tile plot representing the fraction of time spent moving as a colour intensity.
			Each individual is represented by a row and time, on the x-axis, is binned in 30 minutes.
			A: Uncurated raw data.
			B: Data after the curation step. Time was trimmed and data from dead animals removed. 
			Red `\textbf{+}' symbols show animals that were removed from the subsequent analysis as they had less than five complete days in LL.}
	\label{fig:fig-3}
\end{figure}

@@

The activity of dead or escaped animals is falsely scored as long series of zeros (see, for instance, individual 30 and 18 in Fig~\ref{fig:fig-3}A).
Our \texttt{sleepr} package offers a tool to detect and remove such artefactual data.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# remove data after death}
\hlstd{dt} \hlkwb{<-} \hlstd{sleepr}\hlopt{::}\hlkwd{curate_dead_animals}\hlstd{(dt, moving)}
\end{alltt}
\end{kframe}
\end{knitrout}

In addition, we can trim our data to have the same number of days across experiments and individuals.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{dt} \hlkwb{<-} \hlstd{dt[t} \hlopt{%between%} \hlkwd{days}\hlstd{(}\hlkwd{c}\hlstd{(}\hlopt{-}\hlnum{2}\hlstd{,} \hlnum{6}\hlstd{))]}
\hlcom{# same as above}
\hlstd{fig3B} \hlkwb{<-} \hlkwd{ggetho}\hlstd{(dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= label,} \hlkwc{z} \hlstd{= moving))} \hlopt{+}
    \hlkwd{stat_tile_etho}\hlstd{()} \hlopt{+}
    \hlkwd{stat_ld_annotations}\hlstd{(}\hlkwc{x_limits} \hlstd{=} \hlkwd{c}\hlstd{(dt[,} \hlkwd{min}\hlstd{(t)],} \hlnum{0}\hlstd{))} \hlopt{+}
    \hlkwd{stat_ld_annotations}\hlstd{(}\hlkwc{x_limits} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{, dt[,} \hlkwd{max}\hlstd{(t)]),}
                        \hlkwc{ld_colours} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"white"}\hlstd{,} \hlstr{"grey"}\hlstd{))}
\end{alltt}
\end{kframe}
\end{knitrout}

For the purpose of this example, we also exclude animals that died prematurely, and keep only individuals that have \emph{at least five days in LL}. An overview of the curate data can be visualised in Fig~\ref{fig:fig-3}B.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# for each id, we check for validity}
\hlstd{valid_dt} \hlkwb{<-} \hlstd{dt[,} \hlkwd{.}\hlstd{(}\hlkwc{valid} \hlstd{=} \hlkwd{max}\hlstd{(t)} \hlopt{>} \hlkwd{days}\hlstd{(}\hlnum{5}\hlstd{)),} \hlkwc{by} \hlstd{= id]}
\hlcom{# a vector of all valid ids}
\hlstd{valid_ids} \hlkwb{<-} \hlstd{valid_dt[valid} \hlopt{==} \hlstd{T, id]}
\hlcom{# filter dt with the valid ids}
\hlstd{dt} \hlkwb{<-} \hlstd{dt[id} \hlopt{%in%} \hlstd{valid_ids]}
\hlkwd{summary}\hlstd{(dt)}
\end{alltt}
\begin{verbatim}
## behavr table with:
##  52	individuals
##  9	metavariables
##  3	variables
##  1.19546e+05	measurements
##  1	key (id)
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\subsection*{Double-plotted actograms}
``Double-plotted actograms'' are a common visualisation of periodicity and rhythmicity in circadian experiments.
In  \nameref{S1-Fig}A, we show the double-plotted actograms of each animal.
A representative sample of four individuals for each genotype is shown in Fig~\ref{fig:fig-4}A.

@@

\subsection*{Periodograms}
Ultimately, in order to quantify periodicity and rhythmicity, we compute periodograms.
Several methods are implemented in \texttt{zeitbebr} ($\chi{}^2$, Lomb-Scargle and autocorrelation).
In this example, we generate $\chi{}^2$ periodograms and lay them out in a grid.
Periodograms dor the subset of eight animals used in Fig~\ref{fig:fig-4}A is shown in Fig~\ref{fig:fig-4}B. 
See \nameref{S1-Fig}B for the visualisation of all individuals.

@@

\begin{figure}[!h]
	%\includegraphics[width=1\textwidth]{fig/fig-4.pdf}
	\caption{{\bf Visualisation of the periodicity in activity of eight representative animals.}
		A: Double plotted actograms showing activity over the experiment. Transition from LD to LL happens at day 0.
		B: $\chi{}^2$ periodograms over the LL part of the experiment matching the animals in A.
		The blue cross represents the first peak (if present) above the significance threshold (red line).
		Titles on top of each facet refer to the label allocated to each individual.
		See \nameref{S1-Fig} for all 52 animals.
	}
	\label{fig:fig-4}
\end{figure}

\subsection*{Population statistics}

As shown in the original study, double-plotted actograms and periodograms suggest that NKCCOX/+ flies are mostly 
arhythmic in LL whilst Tim/NKCCOX appear to have a consistent, long-period rhythm.
To visualise this at the population scale, we can plot an average periodogram (see Fig~\ref{fig:fig-5}A):

@@

\begin{figure}[!h]
	%\includegraphics[width=1\textwidth]{fig/fig-5.pdf}
	\caption{{\bf Population statistics on circadian phenotype.}
			A: Average periodograms. 
			      The aggregated relative power of the periodogram of all animals.
			      The solid lines and the shaded areas show population means and their 95\% bootstrap confidence interval, respectively.
			B: Frequencies of rhythmic animals.
			      Number of rhythmic animals (\emph{i.e.} with a significant peak) in each genotypes.
			      Dark and clear fillings indicate rhythmic and arhythmic animals, respectively.
      C: Peak periodicity power and average.
			      Values of the peak period for animals with a significant peak (\emph{i.e.} rhythmic).
			      Individual animals are shown by dots whose size represent relative power of the peak period.
			      The error bars are 95\% bootstrap confidence interval on the population mean.
			      }
	\label{fig:fig-5}
\end{figure}

To further quantify this difference, we can show the number of rhythmic animals -- \emph{i.e.} individuals for which a peak was found -- in each groups (see Fig~\ref{fig:fig-5}B).
Then, we can compare the average value of the peak for the rhythmic animals (see Fig~\ref{fig:fig-5}C).
First of all, we compute a summary per individual (\texttt{by=id}):

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{summary_dt} \hlkwb{<-} \hlstd{per_dt[,}\hlkwd{.}\hlstd{(}
                        \hlkwc{first_peak_period} \hlstd{= period[peak}\hlopt{==}\hlnum{1}\hlstd{],}
                        \hlkwc{first_peak_rel_power} \hlstd{= power[peak}\hlopt{==}\hlnum{1}\hlstd{]} \hlopt{-} \hlstd{signif_threshold[peak}\hlopt{==}\hlnum{1}\hlstd{],}
                        \hlkwc{is_rhythmic} \hlstd{=} \hlkwd{any}\hlstd{(peak}\hlopt{==}\hlnum{1}\hlstd{)),}
                      \hlkwc{by}\hlstd{=id]}
\hlcom{# rejoin metadata}
\hlstd{summary_dt} \hlkwb{<-} \hlkwd{rejoin}\hlstd{(summary_dt)}
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# standard ggplot}
\hlstd{fig5B} \hlkwb{<-} \hlkwd{ggplot}\hlstd{(summary_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{x} \hlstd{= genotype,}
                                \hlkwc{fill} \hlstd{= genotype,}
                                \hlkwc{alpha} \hlstd{= is_rhythmic}
                                \hlstd{))} \hlopt{+}
              \hlkwd{geom_bar}\hlstd{(}\hlkwc{colour}\hlstd{=}\hlstr{"black"}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# standard ggplot}
\hlstd{fig5C} \hlkwb{<-} \hlkwd{ggplot}\hlstd{(summary_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= first_peak_period,}
                                \hlkwc{x} \hlstd{= genotype))} \hlopt{+}
              \hlcom{# draw the mean of each genotype group}
              \hlkwd{stat_summary}\hlstd{(}\hlkwc{fun.y} \hlstd{= mean,} \hlkwc{geom} \hlstd{=} \hlstr{"point"}\hlstd{,} \hlkwc{shape}\hlstd{=}\hlnum{3}\hlstd{)} \hlopt{+}
              \hlcom{# draw bootstrap confidence intervals}
              \hlkwd{stat_summary}\hlstd{(}\hlkwc{fun.data} \hlstd{= mean_cl_boot,} \hlkwc{geom} \hlstd{=} \hlstr{"errorbar"}\hlstd{)} \hlopt{+}
              \hlcom{# shows all individuals as points. }
              \hlcom{# the size of the point expresses the power of the peak}
              \hlkwd{geom_jitter}\hlstd{(}\hlkwd{aes}\hlstd{(}\hlkwc{colour}\hlstd{= genotype,}
                              \hlkwc{size}\hlstd{=first_peak_rel_power),}
                              \hlkwc{alpha}\hlstd{=}\hlnum{.67}\hlstd{)} \hlopt{+}
              \hlcom{# We would like to convert time in hour}
              \hlkwd{scale_y_hours}\hlstd{(}\hlstr{"Period"}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

\texttt{R} provides one of the richest statistical toolbox available, which allows users to go deeper tin he analysis of the extracted variables.
One could, for instance, perform a $\chi{}^2$ test on the number of rhythmic \emph{vs} arhythmic flies in both genotypes.
To address the same question, we fit a binomial generalised linear model:

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{fit} \hlkwb{<-} \hlkwd{glm}\hlstd{(is_rhythmic} \hlopt{~} \hlstd{genotype, summary_dt,} \hlkwc{family}\hlstd{=}\hlstr{"binomial"}\hlstd{)}

@@

The result shows a strong positive effect of genotype Tim/NKCCOX on the probability of being rhythmic ($p$-value $6.27 \times 10^{-06}$):

Lastly, we can generate a table that compute arbitrary population statistics for each genotype:
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{result_dt} \hlkwb{<-} \hlstd{summary_dt[,}
          \hlkwd{.}\hlstd{(}
            \hlkwc{mean_period} \hlstd{=} \hlkwd{mean}\hlstd{(first_peak_period,} \hlkwc{na.rm} \hlstd{= T)} \hlopt{/} \hlkwd{hours}\hlstd{(}\hlnum{1}\hlstd{),}
            \hlkwc{sd_period} \hlstd{=} \hlkwd{sd}\hlstd{(first_peak_period,} \hlkwc{na.rm} \hlstd{= T)} \hlopt{/} \hlkwd{hours}\hlstd{(}\hlnum{1}\hlstd{),}
            \hlkwc{n_rhythmic} \hlstd{=} \hlkwd{sum}\hlstd{(is_rhythmic),}
            \hlkwc{n} \hlstd{= .N,}
            \hlkwc{percent_rhythmic} \hlstd{=} \hlnum{100} \hlopt{*} \hlkwd{sum}\hlstd{(is_rhythmic)}\hlopt{/}\hlstd{.N}
            \hlstd{),}
          \hlkwc{by} \hlstd{= genotype}
          \hlstd{]}
\hlstd{result_dt}
\end{alltt}
\begin{verbatim}
##      genotype mean_period sd_period n_rhythmic  n percent_rhythmic
## 1:   NKCCOX/+    25.22500  3.598495          4 22         18.18182
## 2: Tim/NKCCOX    26.21429  2.363568         28 30         93.33333
\end{verbatim}
\end{kframe}
\end{knitrout}

This examples shows how \texttt{rethomics} can be used from the raw data to making publication quality figures and statistics.
We were able to comprehensively analyse classical circadian data with a few line of code.
This workflow applies  particularly really well to much larger data sets and provides a large degree of flexibility taht will allow
biologists to tune their analysis to their specific questions.
